---
title: "Final Project Report"
author: "Jamie Valencia Lopez, Jonathan Wu, Faith Ang, Melody Lu"
date: "2026-03-09"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
library(dplyr)
library(ggplot2)
library(patchwork)
library(broom)
library(boot)
library(themis)
library(recipes)

# Set the random seed for reproducibility
set.seed(167)
```

# Introduction

## Overall Objective

What causes airplane delays and cancellations? Airplane delays and cancellations could be due to various factors. However, the disruptions are complex and poorly understood, even though those problems create significant costs and inconvenience for all passengers, airlines, and airports involved. 

This project aims to identify the key factors that influence the flight delays, cancellations, and on-time performance using the `nycflights13` dataset. Our group seeks to examine the conditions and factors, such as weather, time of day, and airport, that contribute to the most delays and cancellations. We are also developing models that predict whether a plane will be delayed and how delayed, or canceled. Through this analysis, we hope to gain a better understanding of patterns in flight delays and reliability as well as to provide insights that could improve flight efficiency and experience.

## Importance and Interest
The problem is important because flight disruptions create significant costs and inconvenience for all passengers. It is interesting to explore how different variables like weather and carrier operations intersect to create these complex patterns in a major transit hub like New York City.

## Research Questions
1. **What conditions cause the most delays?**
2. **What are the optimal conditions for an on-time departure?**
3. **Which airlines have the most delays?**
4. **Which airlines take off while others don’t in cases of bad weather/not optimal conditions?**
5. **What airports have the best on-time rate? (On-time flights / all flights). What do these airports have in common?**
6. **Can we predict if a plane will be delayed?**
7. **Can we predict how delayed a plane will be?**
8. **Can we predict if a flight will be cancelled based on weather conditions? What are the most important variables to predict?**

## Goals and Hypotheses
* To identify key environmental and temporal factors influencing flight performance.
* To compare airline resilience during non-optimal conditions.
* To build predictive models for delay duration and cancellation probability.


# Data Description

Dataset: https://nycflights13.tidyverse.org/

The dataset that we are using is the "nycflights13" dataset from tidyverse. This dataset contains all flight information on all flights that departed from New York City in 2013 from the airports EWR, JFK, and LGA to places inside the United States, Puerto Rico, and the American Virgin Islands.The dataset is broken up into 5 data tables. In total there are 336,776 flights in the dataset stored in the flights data table. The weather data table contains hourly information for each airport on weather. The planes data table contains information about make and model of the planes, airport contains location information about each airport, and the airline data table contains information about abbreviations for each airline. We plan to merge these tables based on variables such as carrier, tailnum, origin, and time_hour, which are seen across tables. By joining these, we can create a complete dataset that allows us to analyze how specific weather conditions, aircraft age, and airport characteristics simultaneous impact flight punctuality and cancellations,.


## Data Overview

Here is a summary of the data. The variables that we will be using are mostly going to be in the weather and flights dataset. In the flights dataset we will be using scheduled departure time, actual departure time, tail number, origin, and departure delay. From the weather data table we will be using origin, month, humidity, wind direction, wind speed, precipitation, visibility, and time_hour. Those variables will help us figure out the weather conditions that are correlated with the departure delayed flights.

```{r, echo = F}

summary(flights)

```



```{r}
colSums(is.na(flights)) #checking for N/A values in variables
```

The flight data table does have some NA values in the departure time, delay, and arrival time due to flights that ended up being cancelled. To sort those out we will create a new binary column to indicate whether the flight was cancelled or not. The data initially appears to be clean as the data has a uniform format. The two datasets will also be kept separate as combining the flight information and weather information into one dataset would make accessing more complicated. The dataset is sufficient to answer the questions as it contains the relevant information related to our questions.

***

\newpage

# EDA

## Early Exploratory Data Analysis

The first thing that needs to be done is to clean the data of the NA values. This ensures that the data will cause minimal issues when working with it. In addition, a new column is added called ```status``` which will declare which flights were delayed or on time, and the subsequent boxplot separated by carrier, which will help determine that if the flight was delayed, what the spread of delay looks like. This can provide insights on average delay, and if there are very large outliers.

```{r}

hour_plot_data <- flights %>%
     mutate(status = if_else(dep_delay <= 0, "On-time", "Delayed")) %>% #status
     filter(!is.na(status))

```


This histogram separates the time of the departure delay by carrier, and will help in seeing which carriers has the most delays be seeing the spread of the plots

```{r, echo = F}

flights_delayed <- filter(hour_plot_data, status == "Delayed")

ggplot(data = flights_delayed, aes(x = carrier, y = dep_delay, fill = carrier)) +
  geom_boxplot(outlier.shape = NA) +
  coord_cartesian(ylim = c(0, 250))

```
\newpage

These histograms use ```hour_plot_data```, and counts how many flights there are in the data set, sorted by carriers. This helps with getting the idea with the spread of the total flights that each carrier had.

```{r, fig.height= 7, echo = F}
#using hour_plot_data to see how many plane trips belonged to each carrier
ggplot(data = hour_plot_data) +
  geom_bar(mapping = aes(x = carrier, fill = carrier)) +
  geom_text(stat = "count", aes(x = carrier, label = after_stat(count)), vjust = -1)
```

\newpage

This same boxplot data, but further separated into “on time” and “delayed” can help answer the question of which airlines are more delayed than others, and a visual guide into seeing the distribution of “on time” versus “delayed” flights.

```{r, fig.height= 7, echo = F}
#using hour_plot_data to see bargraph of planes that were on time or delayed
ggplot(data = hour_plot_data) +
  geom_bar(mapping = aes(x = status, fill = carrier)) +
  geom_text(stat = "count", aes(x = status, label = after_stat(count)), vjust = 0, size = 3) +
  facet_wrap(~ carrier, nrow = 2)
```

***
\newpage

## Data analysis for Research Questions

1. **What conditions cause the most delays?**

We were curious to ask the first question "**What conditions cause the most delays?**" so we joined the data of weather and flights together into data ```weather_impact```. These graphs are plotting delays and weather conditions to find spikes/critical values. 

```{r, warning = F, echo=F}
weather_impact <- flights %>%
  filter(!is.na(dep_delay)) %>%
  # Selecting only necessary columns to save memory
  select(origin, year, month, day, hour, dep_delay) %>%
  inner_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  # Filter out extreme outliers for a clearer 'tipping point' view
  filter(dep_delay > -20 & dep_delay < 200)

weather_impact %>%
  # 2. Reshape it first so we have a 'metric' column
  pivot_longer(cols = c(visib, wind_speed, precip, pressure), 
               names_to = "metric", 
               values_to = "value") %>%
  # 3. Now we can filter for the factors we want to see
  filter(metric %in% c("visib", "wind_speed", "precip", "pressure")) %>%
  # 4. Plot
  ggplot(aes(x = value, y = dep_delay)) +
  geom_jitter(alpha = 0.02, color = "gray50") + 
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), color = "firebrick") +
  facet_wrap(~metric, scales = "free_x") +
  theme_minimal() +
  labs(title = "Final Weather Threshold Analysis",
       x = "Weather Reading",
       y = "Departure Delay (min)")
```

\newpage

There are a lot of data points so the graphs are hard to read and understand, so we will resort to binning in order to make it easier to understand.

```{r, echo = F, warning=FALSE}

weather_impact %>%
  pivot_longer(cols = c(visib, wind_speed, precip, pressure), 
               names_to = "metric", 
               values_to = "value") %>%
  # Group by the weather metric and round the value to create "bins"
  group_by(metric, value = round(value, 1)) %>% 
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  filter(metric %in% c("visib", "wind_speed", "precip", "pressure")) %>%
  ggplot(aes(x = value, y = avg_delay)) +
  geom_line(color = "gray70") +
  geom_point(aes(color = avg_delay), size = 2) +
  geom_smooth(method = "gam", color = "firebrick", se = FALSE) +
  facet_wrap(~metric, scales = "free_x") +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Average Delay by Weather Intensity",
       subtitle = "Binning data makes the 'tipping point' spikes much more visible",
       y = "Mean Departure Delay (min)")
```
\newpage

The precipitation graph has weird and random trends due to low volume of flights with precipitation higher than .5 inch/hr. This is further cleaned so that the graph stabilizes.

```{r, echo = F}
precip_counts <- flights %>%
  inner_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  filter(precip > 0) %>%
  # Create the same bins used in the histogram (0.02 inches)
  mutate(precip_bin = round(precip / 0.02) * 0.02) %>%
  group_by(precip_bin) %>%
  summarise(n = n()) %>%
  # 2. Filter out bins with 0 flights to fix the 'negative' log issue
  filter(n > 0)

# 3. Plot using geom_col (bars) instead of geom_histogram
ggplot(precip_counts, aes(x = precip_bin, y = n)) +
  geom_col(fill = "darkorange", color = "white", width = 0.02) +
  scale_y_log10() +
  theme_minimal() +
  labs(
    title = "Cleaned Flight Frequency during Precipitation",
    subtitle = "Empty bins removed to stabilize the log scale",
    x = "Precipitation (inches)",
    y = "Number of Flights (Log10)"
  )
```
Importance

This helps to find the threshold, or the kind of weather that is more likely for a plane to be delayed. This contributes to the later question on if it is possible to predict a plane to be delayed based on weather factors.

\newpage

2. **What are the optimal conditions for an on-time departure?**

Below, we can determine that the first hours of the day are best for on-time flights, while the later hours of the day demonstrate higher rates of delayed flights.

```{r, echo = F, message=FALSE, warning=FALSE}
hour_plot_data <- flights |>
     mutate(status = if_else(dep_delay <= 15, "On-time", "Delayed")) |>
     filter(!is.na(status))
 
p1 <- ggplot(hour_plot_data, aes(x = hour, fill = status)) +
     geom_density(alpha = 0.5)
```

Now we will dive into and attempt to answer this question fully in order to complete our EDA. We will look to see what variables we can analyze in the above manner.

```{r}
names(flights)
```

```{r, echo = F, message=FALSE, warning=FALSE}
 p2 <- ggplot(hour_plot_data, aes(x = air_time, fill = status)) +
     geom_density(alpha = 0.5) +
     coord_cartesian(xlim = c(0, 400))
```

```{r, echo = F, message=FALSE, warning=FALSE}
 p3 <- ggplot(hour_plot_data, aes(x = distance, fill = status)) +
     geom_density(alpha = 0.5) +
     coord_cartesian(xlim = c(0, 3000))
```

```{r, echo = F, message=FALSE, warning=FALSE}
p4 <- ggplot(hour_plot_data, aes(x = minute, fill = status)) +
     geom_density(alpha = 0.5)
```

```{r, warning = F}
(p1 + p2) / (p3 + p4)
```

These graphs demonstrate the biggest disparity we see: hour of the day. It demonstrates that flights on-time flights are distributed more heavily on the earlier hours of the day; whereas delayed fllights seem to be centered on the later hours of the day. The rest of the variables (air time, distance and minute) do not seem to show a strong relationship with flight delays.

```{r, warning = F}
combined_data <- flights %>%
  left_join(weather, by = c("origin", "time_hour"))

combined_data <- combined_data %>%
  mutate(across(where(is.character), as.factor))

names(combined_data)
```

```{r, echo = F, message=FALSE, warning=FALSE}
hour_plot_data_combined <- combined_data |>
     mutate(status = if_else(dep_delay <= 15, "On-time", "Delayed")) |>
     filter(!is.na(status))


p5 <- ggplot(hour_plot_data_combined, aes(x = humid, fill = status)) +
     geom_density(alpha = 0.5)

p6 <- ggplot(hour_plot_data_combined, aes(x = pressure, fill = status)) +
     geom_density(alpha = 0.5)

p7 <- ggplot(hour_plot_data_combined, aes(x = visib, fill = status)) +
     geom_density(alpha = 0.5) +
     coord_cartesian(xlim = c(9, 10))

p8 <- ggplot(hour_plot_data_combined, aes(x = wind_dir, fill = status)) +
     geom_density(alpha = 0.5)

p9 <- ggplot(hour_plot_data_combined, aes(x = temp, fill = status)) +
     geom_density(alpha = 0.5)

p10 <- ggplot(hour_plot_data_combined, aes(x = wind_gust, fill = status)) +
     geom_density(alpha = 0.5) +
     coord_cartesian(xlim = c(17,45))

p11 <- ggplot(hour_plot_data_combined, aes(x = dewp, fill = status)) +
     geom_density(alpha = 0.5)

p12 <- ggplot(hour_plot_data_combined, aes(x = precip, fill = status)) +
     geom_density(alpha = 0.5) + 
     coord_cartesian(xlim = c(0,0.015))
```

```{r, warning = F}
(p5 + p6) / (p7 + p8) 
(p9 + p10) / (p11 + p12 )
```
Here we can see that altough there is not a relationship as strong as that seen for hour, variables like dewp and temp demonstrate similar trends. On-time flights seem to be more common for lower temperatures, as well as for lower dew points. We can also derive a similar relationship for precipitation. and humidity, which makes sense intuitively. Pressure and visibility have the opposite relationship, with lower values signifiying a higher concentration of delayed flights.

```{r, warning = F}
planes_clean <- planes %>%
  rename(year_built = year) %>%
  mutate(across(where(is.character), as.factor)) 

master_data <- flights %>%
  left_join(weather, by = c("origin", "time_hour")) %>%
  left_join(planes_clean, by = "tailnum")

names(master_data)
```

```{r, echo = F, message=FALSE, warning=FALSE}
hour_plot_data_combined2 <- master_data |>
     mutate(status = if_else(dep_delay <= 15, "On-time", "Delayed")) |>
     filter(!is.na(status))


p14 <- ggplot(hour_plot_data_combined2, aes(x = seats, fill = status)) +
     geom_density(alpha = 0.5)

p15 <- ggplot(hour_plot_data_combined2, aes(x = speed, fill = status)) +
     geom_density(alpha = 0.5) +
     coord_cartesian(xlim = c(90,250))
```

```{r, warning = F}
(p14 / p15)
```

In these last plots, we see a slight relationship with speed of planes and delays. Lower speeds seem to have a slighty higher concentration of on-time flights. 

Next, we will analyze the relationship of two variable in regards to delayed and on-time flights.

```{r, warning = F}
ggplot(hour_plot_data_combined, aes(x = hour.x, fill = status)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~month.x)
```

```{r, warning = F}
ggplot(hour_plot_data_combined, aes(x = humid, fill = status)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~origin)
```

```{r, warning = F}
hour_plot_data_combined2 |> 
  filter(!is.na(engine)) |> 
  ggplot(aes(x = hour.x, fill = status)) +
    geom_density(alpha = 0.5) +
    facet_wrap(~engine)
```

```{r, warning = F}
ggplot(hour_plot_data_combined2, aes(x = hour.x, fill = status)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~origin)
```

```{r, warning = F}
hour_plot_data_combined2 %>%
  mutate(humid_bin = cut(humid, breaks = seq(0, 100, by = 20))) %>%
  filter(!is.na(humid_bin)) %>%
  ggplot(aes(x = humid_bin, y = dep_delay, fill = origin)) +
  geom_boxplot(outlier.shape = NA) + 
  coord_cartesian(ylim = c(-10, 40)) + 
  facet_wrap(~origin) +
  theme_light() +
  labs(title = "Delay Ranges by Humidity and Origin",
       x = "Humidity Bin", 
       y = "Delay (Mins)") +
  # --- SLANTING LOGIC HERE ---
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
```

**Importance**: Looking at this, there is a certain trend where there are relationships between various variables, and whether or not a flight is delayed. This is the EDA which will be useful in creating the models for our Forward Selection Model

\newpage

3. **Which airlines have the most delays?**

The first thing done is to create a boxplot of the dep_delay to see the spread of delay among airlines like from the initial EDA. However, I found that there are a lot of outliers that caused the boxplots to be skewed.

Going through a different approach, The number of total flights per carrier were counted.

```{r, fig.height= 7, echo = F}
#using hour_plot_data to see how many plane trips belonged to each carrier
ggplot(data = hour_plot_data) +
  geom_bar(mapping = aes(x = carrier, fill = carrier)) +
  geom_text(stat = "count", aes(x = carrier, label = after_stat(count)), vjust = -1)
```


From there, these flights were further separated by if they were on time, or delayed, as shown below.

```{r, fig.height= 7, echo = F}
#using hour_plot_data to see bargraph of planes that were on time or delayed
ggplot(data = hour_plot_data) +
  geom_bar(mapping = aes(x = status, fill = carrier)) +
  geom_text(stat = "count", aes(x = status, label = after_stat(count)), vjust = 0, size = 3) +
  facet_wrap(~ carrier, nrow = 2) +
   labs(
    title = "Status of Flights by Carrier",,
    y = "Flights"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

Just for viewing, another visualization with only the number of delayed fights per carrier was made.

```{r, fig.height= 7, echo = F}
#using hour_plot_data to see bargraph of carriers that were most delayed
ggplot(data = flights_delayed) +
  geom_bar(mapping = aes(x = carrier, color = carrier, fill = carrier)) +
  geom_text(stat = "count", aes(x = carrier, label = after_stat(count)), vjust = 0) +
   labs(
    title = "Number of Delayed Flights by Carrier",
    x = "Carrier",
    y = "Flights Delayed"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
  
```

Above, it can be seen that of all of the carriers, UA appears to have the most delayed flights across all of the carriers. However, this does not necessarily mean that the carrier is the carrier with the most delays. This is because some carriers only have a population that is not as big as the other carriers. So the next idea is to measure the proportion of delayed flights across the different carriers. This was done using a tibble.


```{r, echo = F, message=FALSE, warning=FALSE}

flight_proportions <- flights %>%
  filter(!is.na(dep_delay)) %>% # Remove flights with missing delay info (likely cancelled)
  group_by(carrier) %>%
  summarise(
    total_flights = n(),
    delayed_flights = sum(dep_delay > 0, na.rm = TRUE), # Count delayed flights
    proportion_delayed = delayed_flights / total_flights # Calculate the proportion
  ) %>%
  arrange(desc(proportion_delayed)) # Sort by proportion_delayed in descending order

# View the result
print(flight_proportions)
```

After this, it can be seen that the carrier with the highest rate of delayed flights is not UA, but is actually WN. This can be further seen by visualization of the tibble.

```{r, echo = F, message=FALSE, warning=FALSE, fig.height= 7}

# Plot the proportion of delayed flights by carrier
ggplot(data = flight_proportions, aes(x = carrier, y = proportion_delayed)) +
  geom_bar(stat = "identity", fill = "lightblue") + # Use 'identity' stat as y-values are pre-calculated
  labs(
    title = "Proportion of Delayed Flights by Carrier",
    x = "Carrier",
    y = "Proportion Delayed"
  ) +
  geom_text(aes(label = sprintf("%.2f", proportion_delayed))) +
  scale_y_continuous(labels = scales::percent) + # Format y-axis as percentage
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) # Center the plot title


```

This visualization below is ordered from highest proportion to lowest proportion.

```{r, echo = F, message=FALSE, warning=FALSE, fig.height= 7}

# Plot the proportion of delayed flights by carrier
ggplot(data = flight_proportions, aes(x = reorder(carrier, -proportion_delayed), y = proportion_delayed)) +
  geom_bar(stat = "identity", fill = "lightblue") + # Use 'identity' stat as y-values are pre-calculated
  labs(
    title = "Proportion of Delayed Flights by Carrier",
    x = "Carrier",
    y = "Proportion Delayed"
  ) +
  geom_text(aes(label = sprintf("%.2f", proportion_delayed))) +
  scale_y_continuous(labels = scales::percent) + # Format y-axis as percentage
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) # Center the plot title


```

**Importance**: The importance of this is to help predict and give insight on which carriers tend to be delayed. This is added with possible predictions to see if a flight will be delayed depending on the carrier the flight is from. 

\newpage

4. **Which airlines take off while others don’t in cases of bad weather/ not optimal conditions?**

"Extreme" Weather Definition:

low visibility: `visib` < 3 

strong wind: `wind_speed` > 20

decent rain: `precip` > 0.1


```{r, echo = F}
flights_weather <- flights %>%
  left_join(weather, 
            by = c("year", "month", "day", "hour", "origin"))

flights_weather <- flights_weather %>%
  mutate(weather_condition = case_when(
    visib < 3 | wind_speed > 20 | precip > 0.1 ~ "Extreme",
    TRUE ~ "Normal"
  ))

head(flights_weather)
```


flight counts in extreme weather

```{r, echo = F}
flights_weather %>%
  filter(weather_condition == "Extreme") %>%
  count(carrier) %>%
  arrange(n)
```


**On-time rate per airline**

Here I have on-time defined as 15 minutes or less departure delay as my threshold.

```{r}

airline_resilience <- flights_weather %>%
  filter(!is.na(dep_delay), !is.na(weather_condition)) %>%
  mutate(on_time = dep_delay <= 15) %>%
  group_by(carrier, weather_condition) %>%
  summarise(
    total_flights  = n(),
    on_time_rate   = mean(on_time) * 100,
    .groups = "drop"
  ) %>%
  left_join(airlines, by = "carrier") %>%           # add full airline names
  mutate(name = str_remove(name, " Inc\\.| Co\\.| Airways| Airlines| Air Lines"))

head(airline_resilience)
```

Additionally, I have minimum flights threshold as well to get a better fit.

```{r}
min_flights <- 100  

reliable_carriers <- flights_weather %>%
  filter(weather_condition == "Extreme") %>%
  count(carrier) %>%
  filter(n >= min_flights) %>%
  pull(carrier)

reliable_carriers

airline_resilience <- airline_resilience %>%
  filter(carrier %in% reliable_carriers)

airline_resilience
```


**Normal and Extreme Weather On-Time Rate**

Resilience drop is calculate as normal on-time rate minus extreme on-time rate.

The lower the difference, the more resilient the airline is.

```{r}
resilience_drop <- airline_resilience %>%
  select(carrier, name, weather_condition, on_time_rate) %>%
  pivot_wider(names_from = weather_condition, values_from = on_time_rate) %>%
  mutate(drop = Normal - Extreme) %>%
  arrange(drop)

head(resilience_drop)
```

**Resilient airline ranking**

```{r, collapse=TRUE}
carrier_order <- airline_resilience %>%
  filter(weather_condition == "Extreme") %>%
  arrange(desc(on_time_rate)) %>%
  pull(carrier)

head(carrier_order)

airline_resilience <- airline_resilience %>%
  mutate(carrier = factor(carrier, levels = carrier_order))

head(airline_resilience)
```

**Bar Chart**

```{r, fig.width=10, echo = F}
ggplot(airline_resilience, aes (x = carrier, y = on_time_rate, fill = weather_condition)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.5) +
  geom_text(
    aes(label = sprintf("%.0f%%", on_time_rate)),
    position = position_dodge(width = 0.75),
    vjust = -0.4, size = 2.8, fontface = "bold") +
  facet_wrap(~ weather_condition, ncol = 1, scales = "free_y") +
  scale_fill_manual(
    values = c("Normal" = "#4A90D9", "Extreme" = "#E8523A"),
    name   = "Weather Condition") +
  scale_y_continuous(labels = scales::label_percent(scale = 1),
                     limits = c(0, 105), expand = c(0, 0)) +
  labs(title    = "Airline On-Time Departure Rate: Normal vs. Extreme Weather",
       subtitle = "Extreme = visibility < 3 mi, wind > 20 mph, or precipitation > 0.1 in/hr\nOn-time defined as dep_delay ≤ 15 min\nCarriers arranged by Extreme-weather on-time rate (left = most resilient)",
       x = "Carrier", y = "On-Time Rate (%)",
       caption  = "Source: nycflights13") +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 10, color = "grey40"),
        strip.text = element_text(face = "bold", size = 11),
        legend.position = "none")
```
Importance: This gives an idea on the rate of departure based on the weather, and notice that there are less rates of on-time departure when the weather is considered extreme.


\newpage

# Modeling

## Model 1: Forward Selection Model with RSS and Polynomial Regression
- Model specification
- Interpretation of coefficients

Here, we will implement a linear regression model to predict the amount of delay a flight will have (arr_delay). We use a custom forward selection strategy to identify key predictors and evaluate the model using 10-fold cross-validation.

Here we create the functions to calculate the MSE, as well as forward selection with lowest RSS. We are working with the weather, planes and flights tables.


```{r}
# Function to calculate Mean Squared Error (MSE)
calculateMSE <- function(y, yhat) {
  residuals <- y - yhat
  mse <- mean(residuals^2)
  return(mse)
}

# Automated Forward Selection Function
# This function picks the top 'n_vars' predictors based on the lowest RSS.
forward_selection <- function(data, response, n_vars) {
  all_predictors <- names(data)[names(data) != response]
  selected_vars <- c()
  
  # Iterate one time in the outer loop for each variable
  for (i in 1:n_vars) {
    best_rss <- Inf
    best_var <- NULL
    remaining_vars <- setdiff(all_predictors, selected_vars)
    
    # Inner loop will iterate through all the variables
    for (var in remaining_vars) {
      # Choose the variable that has the lowest RSS value for the model
      formula_vars <- c(selected_vars, var)
      current_formula <- as.formula(paste(response, "~", 
                                    paste(formula_vars, collapse = " + ")))
      
      # Fit model and calculate RSS
      fit <- lm(current_formula, data = data)
      current_rss <- sum(resid(fit)^2)
      
      # Update the best variable for this iteration
      if (current_rss < best_rss) {
        best_rss <- current_rss
        best_var <- var
      }
    }
    selected_vars <- c(selected_vars, best_var)
    cat("Step", i, ": Added", best_var, "(RSS:", round(best_rss, 2), ")\n")
  }
  return(selected_vars)
}
```

```{r, echo = FALSE}
# Join all the tables and clean the data
flights_clean <- flights |>
  # Join the weather table by origin, year, month, day and hour
  inner_join(weather, by = c("origin", "year", "month", "day", "hour")) |>
  # Join the planes table by tailnum
  inner_join(planes, by = "tailnum", suffix = c("_flight", "_plane")) |>
  filter(dep_delay < 240) |> 
  # Add carrier and origin to the select list
  select(dep_delay, hour, temp, visib, wind_speed, precip, 
         seats, distance, origin, carrier, month) |>
  # Convert to factors so the lm() function treats them as categories
  mutate(
    origin = as.factor(origin),
    carrier = as.factor(carrier),
    month = as.factor(month)
  ) |>
  drop_na()
```

Retroactive EDA: Here, we visualize the relationship of the variables we will choose for our model in order to get a better understanding of their relationship with the response variable. Looking at each variable individually, we see that their relationship is not inherently linear, which is why we will need make use of polynomial regression in the future.

```{r, echo = FALSE}
plot_data <- flights_clean |>
  select(dep_delay, hour, month, carrier, visib, precip, wind_speed) |>
  # Convert everything to numeric for the scatter plot (except dep_delay)
  mutate(across(-dep_delay, as.numeric)) |>
  pivot_longer(
    cols = c(hour, month, carrier, visib, precip, wind_speed),
    names_to = "variable",
    values_to = "value"
  )

# Create the single scatter plot
ggplot(plot_data, aes(x = value, y = dep_delay, color = variable)) +
  geom_smooth(alpha = 0.4) + # Alpha helps with overlapping points
  labs(
    title = "Explanatory Variables vs. Departure Delay",
    x = "Variable Value",
    y = "Departure Delay (minutes)",
    color = "Variable Name"
  ) +
  theme_minimal() +
  coord_cartesian(xlim = c(0,25))
```


**Section 1: Forward Selection Model with RSS**

Here, we will call the forward selection model with the top 6 variables to predict departure delay. The adjusted R-squared value begins to plateau after adding more than 6 variables to the model. For now, we will work under the assumption that all the relationships between the variables are linear, and we will work on the optimal degree in the next section.

```{r}
# Run the forward selection to get the top 3 variables
top_vars <- forward_selection(data = flights_clean, 
                             response = "dep_delay", 
                             n_vars = 6)

cat("\nTop selected variables:", paste(top_vars, collapse = ", "), "\n")
```

Here, we can calculate the weighted 10-fold Cross Validation for our model.
```{r, echo = FALSE}
# Setup for 10-Fold CV
formula_cv <- as.formula(paste("dep_delay ~", paste(top_vars, collapse = " + ")))

# 2. Fit the model using glm()
glm.adv <- glm(formula_cv, data = flights_clean)

# 3. Run the 10-fold Cross-Validation
cv.result <- cv.glm(flights_clean, glm.adv, K = 10)

# 4. Extract raw CV estimate (MSE)
cat("cv.glm Weighted CV MSE:", round(cv.result$delta[1], 6), "\n")
```

Finally display the summary of our model.

```{r}
final_formula <- as.formula(paste("dep_delay ~", paste(top_vars, collapse = " + ")))
lm_final <- lm(final_formula, data = flights_clean)

summary(lm_final)
```

We have a pretty low Adjusted R-squared value of 0.09691, meaning that we can only explain a bit lower than 10% of the variance in our response variable. The MSE of 1057.256 means that there is an average error of around 32.5 minutes per prediction. We have established that the relationship between the explanatory variables and the response variable is not necessary linear, therefore, we will implement a model in which we use polynomial regression to augment our model prediction power.

**Section 2: Polynomial Regression GAM Model**

Here, we will use the 10-fold validation model, processing the variables that we have already selected with the best RSS values. We take a look at each variable and test their 10-fold CV MSE value from powers 1 through 5 and select the power that has the lowest MSE value. However, to prevent  overfitting and to create a more genreal model with less bias, we will use the elbow method in order to pick the most optimal degree for our variables. This will allow us to still increase the performance of the model while mantaining some generarily for unseen data.

```{r, echo = FALSE}
# List of numeric variables to test
vars_to_test <- c("hour", "visib", "wind_speed", "precip")
max_poly <- 5

# Create a data frame to store results for plotting
cv_results_df <- data.frame()

for (v in vars_to_test) {
  cat("Processing variable:", v, "\n")
  
  for (d in 1:max_poly) {
    # Dynamically create the formula: dep_delay ~ poly(var, d)
    formula_str <- as.formula(paste("dep_delay ~ poly(", v, ",", d, ")"))
    
    # Fit the GLM
    fit <- glm(formula_str, data = flights_clean)
    
    # Calculate 10-fold CV
    cv_err <- cv.glm(flights_clean, fit, K = 10)$delta[1]
    
    # Save result
    cv_results_df <- rbind(cv_results_df, data.frame(Variable = v, Degree = d, MSE = cv_err))
  }
}

# Find the degree with the minimum MSE for each variable
optimal_values <- cv_results_df %>%
  group_by(Variable) %>%
  filter(MSE == min(MSE)) %>%
  select(Variable, Degree, MSE)

print(optimal_values)
```

```{r, echo = FALSE}
glm.optimal <- glm(dep_delay ~ poly(hour, 3, raw = TRUE) + 
                   month + 
                   carrier + 
                   poly(visib, 3, raw = TRUE) + 
                   poly(precip, 3, raw = TRUE) + 
                   poly(wind_speed, 2, raw = TRUE), 
                   data = flights_clean)

# Perform 10-fold CV to get the MSE
cv.optimal <- cv.glm(flights_clean, glm.optimal, K = 10)

# Output the result
cat("Optimal Model 10-fold CV MSE:", cv.optimal$delta[1], "\n")
```

The optimal model 10-fold CV MSE is 1050.117 , meaning that there is a slight improvement from the previous linear model. It has decreased by around 7 points.

```{r}
lm.optimal <- lm(dep_delay ~ poly(hour, 3, raw = TRUE) + 
                   month + 
                   carrier + 
                   poly(visib, 3, raw = TRUE) + 
                   poly(precip, 3, raw = TRUE) + 
                   poly(wind_speed, 2, raw = TRUE), 
                   data = flights_clean)

summary(lm.optimal)
```

Additionally, we see here that the adjusted R-Squared value has increased to 0.1031 from the previous 0.09691; we can now explain over 10% of the variance in our response variable.


This plot demonstrates how we picked the optimal degree for the model. We choose the degrees based on the elbow method: 3 for `visib`, `hour` and `precip`. 2 for `wind_speed`.
```{r, echo = FALSE}
ggplot(cv_results_df, aes(x = Degree, y = MSE)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  facet_wrap(~Variable, scales = "free_y") +
  theme_minimal() +
  labs(title = "10-Fold CV MSE vs. Polynomial Degree",
       subtitle = "The 'Optimal Degree' is the lowest point on each curve",
       x = "Polynomial Degree",
       y = "Mean Squared Error (MSE)")
```



```{r, echo = FALSE}
# Generate diagnostic plots
plot(lm.optimal)

```

**Model 1 Conclusion**

While we were able to achieve better results while using polynomial multiple regression, we still were only to explain around 10% of the variance with our model. We also note that the residuals indicate that our model was better at interpreting lower department delay numbers. While not perfect, this model can be used to predict general patterns of delay. More data and more complex models might be necessary to produce stronger predictions. There are some underlying relationships between models that could be causing the accuracy of the model to be significantly lower, including collinearity between the weather variables (visibility, precipitation, and wind speed are likely to be related). Doing so can allow us to fit the model with the explicit variable relationships in mind. Additionally, we could build a model that predicts shorter delays. As we see from the residual Q-Q plot, our model is able to predict smaller delays at a much better rate.

## Model 2: Logistic Regression
- Model fitting
- Important predictors

Number of cancelled flights per New York airport
```{r, echo = F}
flights %>%
  filter(is.na(dep_time)) %>% # Focus only on cancellations
  ggplot(aes(x = origin, fill = origin)) +
  geom_bar() +
  labs(
    title = "Total Number of Cancelled Flights by Airport",
    subtitle = "NYC Area Airports (2013)",
    x = "Origin Airport",
    y = "Number of Cancellations"
  ) +
  theme_minimal() +
  guides(fill = "none") # Remove legend since X axis covers it
```

Cancellation probablilty across different weather conditions
```{r, echo = F}
df_weather <- flights %>%
  mutate(is_cancelled = ifelse(is.na(dep_time), 1, 0)) %>%
  inner_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(is_cancelled, temp, wind_speed, precip, visib, humid)

# 2. Reshape and Bin
# We use ntile() to create 10 equally sized groups (deciles) for each variable
df_binned <- df_weather %>%
  pivot_longer(cols = -is_cancelled, names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(variable) %>%
  mutate(bin = ntile(value, 10)) %>% 
  group_by(variable, bin) %>%
  summarise(
    mean_val = mean(value),
    cancel_rate = mean(is_cancelled),
    .groups = 'drop'
  )

# 3. Create the Faceted Plot
ggplot(df_binned, aes(x = mean_val, y = cancel_rate)) +
  geom_line(color = "firebrick", size = 1) +
  geom_point(color = "firebrick", size = 2) +
  facet_wrap(~variable, scales = "free_x", ncol = 2) +
  labs(
    title = "Cancellation Probability Across Weather Deciles",
    subtitle = "Comparing how risk increases as weather conditions change",
    x = "Mean Value of Weather Variable (per bin)",
    y = "Cancellation Rate (Probability)"
  ) +
  theme_minimal()
```

## Model Building

Create a subset of the data where it is just the cancelled flights. 
```{r, echo = F}
weather_flights <- flights %>%
  mutate(is_cancelled = ifelse(is.na(dep_time), 1, 0)) %>%
  inner_join(weather, by = c("origin", "year", "month", "day", "hour"))
weather_clean <- weather_flights %>%
  select(is_cancelled, temp, dewp, humid, wind_speed, precip, pressure, visib) %>%
  drop_na()
```

Initial logistic regression model
```{r, echo = F}
log_mod <- glm(is_cancelled ~ temp + dewp + humid + wind_speed + precip + pressure + visib, 
               data = weather_clean, 
               family = binomial(link = "logit"))

model_results <- tidy(log_mod, conf.int = TRUE) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  arrange(desc(odds_ratio))

print(model_results)
```

Important variables with low p-values: wind_speed, humidity, visibility, pressure
Non-important variables with high p-values: precipitation, dewpoint, temperature
Very high odds ratio indicates not converging

```{r}
summary(log_mod)
```
```{r}
mlr.prob <- predict(log_mod, type = "response")
mlr.class <- ifelse(mlr.prob > 0.15, 1, 0)
mean(weather_clean$is_cancelled != mlr.class)
```
Misclassification rate is at 2.1 percent however only 2 percent of the flights in the dataset were cancelled, the model is the same as not looking at weather data at all

```{r}
cv.glm(weather_clean, log_mod, K = 10)$delta[1]
```
Average Weighted Error Rate of .0971 (Penalty score) Score is slightly better than guessing 

```{r}
improved_probs <- predict(log_mod, type = "response")
improved_preds <- ifelse(improved_probs > 0.15, 1, 0)

conf_table <- table(Predicted = improved_preds, Actual = weather_clean$is_cancelled)

print(conf_table)
```
Model is terrible at predicting cancelled flights. 98% accurate but unable to determine flights that don't take off.

## Improved Model

Creating a new model without the unnecessary variables and with relevant variables scaled.
```{r}
data_to_model <- flights %>%
  mutate(is_cancelled = if_else(is.na(dep_time), 1, 0)) %>%
  # Joining by the keys present in both
  inner_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(is_cancelled, month, hour, temp, humid, wind_speed, precip, pressure, visib) %>%
  drop_na() %>%
  mutate(
    # Explicitly use across() to avoid 'closure' errors
    is_cancelled = as.factor(is_cancelled),
    month = as.factor(as.character(month)),
    hour = as.factor(as.character(hour))
  )
```

Downsizing the data so its a 50/50 split of cancelled and not cancelled flights
```{r}
flight_recipe <- recipe(is_cancelled ~ ., data = data_to_model) %>%
  step_downsample(is_cancelled) %>% 
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  prep()

# Create the balanced training set
balanced_train <- bake(flight_recipe, new_data = NULL)
```

Removed irrelevant variables from new model
```{r, echo = F}
improved_model <- glm(is_cancelled ~ humid + wind_speed + pressure + visib, 
                      data = balanced_train, 
                      family = binomial(link = "logit"))

# Check the results - you should see much more reasonable Odds Ratios now
summary(improved_model)
```
```{r}
model_results <- tidy(improved_model, conf.int = TRUE) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  arrange(desc(odds_ratio))

print(model_results)
```


```{r}
probs <- predict(improved_model, newdata = data_to_model, type = "response")
preds <- as.factor(ifelse(probs > 0.5, 1, 0))

# Create the Confusion Matrix
conf_matrix <- table(Predicted = preds, Actual = data_to_model$is_cancelled)
print(conf_matrix)

# Calculate Sensitivity (Recall)
# (True Positives) / (Total Actual Positives)
recall <- conf_matrix[2,2] / sum(conf_matrix[,2])
print(paste("Recall (Cancellations caught):", round(recall, 4)))
```

```{r, echo = F}
plot(improved_model, 
     which = 2, 
     main = "Q-Q Plot: Model Deviance Residuals") 
```


