---
title: "Question6Model"
output: pdf_document
date: "2026-03-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load necessary libraries
```{r cars}
library(ggplot2)
library(dplyr)
library(nycflights13)
library(broom)
library(tidyverse)
library(boot)
library(themis)
library(recipes)
```

## EDA

Number of cancelled flights per New York airport
```{r}
flights %>%
  filter(is.na(dep_time)) %>% # Focus only on cancellations
  ggplot(aes(x = origin, fill = origin)) +
  geom_bar() +
  labs(
    title = "Total Number of Cancelled Flights by Airport",
    subtitle = "NYC Area Airports (2013)",
    x = "Origin Airport",
    y = "Number of Cancellations"
  ) +
  theme_minimal() +
  guides(fill = "none") # Remove legend since X axis covers it
```

Cancellation probablilty across different weather conditions
```{r}
df_weather <- flights %>%
  mutate(is_cancelled = ifelse(is.na(dep_time), 1, 0)) %>%
  inner_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(is_cancelled, temp, wind_speed, precip, visib, humid)

# 2. Reshape and Bin
# We use ntile() to create 10 equally sized groups (deciles) for each variable
df_binned <- df_weather %>%
  pivot_longer(cols = -is_cancelled, names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(variable) %>%
  mutate(bin = ntile(value, 10)) %>% 
  group_by(variable, bin) %>%
  summarise(
    mean_val = mean(value),
    cancel_rate = mean(is_cancelled),
    .groups = 'drop'
  )

# 3. Create the Faceted Plot
ggplot(df_binned, aes(x = mean_val, y = cancel_rate)) +
  geom_line(color = "firebrick", size = 1) +
  geom_point(color = "firebrick", size = 2) +
  facet_wrap(~variable, scales = "free_x", ncol = 2) +
  labs(
    title = "Cancellation Probability Across Weather Deciles",
    subtitle = "Comparing how risk increases as weather conditions change",
    x = "Mean Value of Weather Variable (per bin)",
    y = "Cancellation Rate (Probability)"
  ) +
  theme_minimal()
```

## Model Building

Create a subset of the data where it is just the cancelled flights. 
```{r}
weather_flights <- flights %>%
  mutate(is_cancelled = ifelse(is.na(dep_time), 1, 0)) %>%
  inner_join(weather, by = c("origin", "year", "month", "day", "hour"))
weather_clean <- weather_flights %>%
  select(is_cancelled, temp, dewp, humid, wind_speed, precip, pressure, visib) %>%
  drop_na()
```

Initial logistic regression model
```{r}
log_mod <- glm(is_cancelled ~ temp + dewp + humid + wind_speed + precip + pressure + visib, 
               data = weather_clean, 
               family = binomial(link = "logit"))

model_results <- tidy(log_mod, conf.int = TRUE) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  arrange(desc(odds_ratio))

print(model_results)
```

Important variables with low p-values: wind_speed, humidity, visibility, pressure
Non-important variables with high p-values: precipitation, dewpoint, temperature
Very high odds ratio indicates not converging

```{r}
summary(log_mod)
```
```{r}
mlr.prob <- predict(log_mod, type = "response")
mlr.class <- ifelse(mlr.prob > 0.15, 1, 0)
mean(weather_clean$is_cancelled != mlr.class)
```
Misclassification rate is at 2.1 percent however only 2 percent of the flights in the dataset were cancelled, the model is the same as not looking at weather data at all

```{r}
cv.glm(weather_clean, log_mod, K = 10)$delta[1]
```
Average Weighted Error Rate of .0971 (Penalty score) Score is slightly better than guessing 

```{r}
improved_probs <- predict(log_mod, type = "response")
improved_preds <- ifelse(improved_probs > 0.15, 1, 0)

conf_table <- table(Predicted = improved_preds, Actual = weather_clean$is_cancelled)

print(conf_table)
```
Model is terrible at predicting cancelled flights. 98% accurate but unable to determine flights that don't take off.

## Improved Model

Creating a new model without the unnecessary variables and with relevant variables scaled.
```{r}
data_to_model <- flights %>%
  mutate(is_cancelled = if_else(is.na(dep_time), 1, 0)) %>%
  # Joining by the keys present in both
  inner_join(weather, by = c("origin", "year", "month", "day", "hour")) %>%
  select(is_cancelled, month, hour, temp, humid, wind_speed, precip, pressure, visib) %>%
  drop_na() %>%
  mutate(
    # Explicitly use across() to avoid 'closure' errors
    is_cancelled = as.factor(is_cancelled),
    month = as.factor(as.character(month)),
    hour = as.factor(as.character(hour))
  )
```

Downsizing the data so its a 50/50 split of cancelled and not cancelled flights
```{r}
flight_recipe <- recipe(is_cancelled ~ ., data = data_to_model) %>%
  step_downsample(is_cancelled) %>% 
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  prep()

# Create the balanced training set
balanced_train <- bake(flight_recipe, new_data = NULL)
```

Removed irrelevant variables from new model
```{r}
improved_model <- glm(is_cancelled ~ humid + wind_speed + pressure + visib, 
                      data = balanced_train, 
                      family = binomial(link = "logit"))

# Check the results - you should see much more reasonable Odds Ratios now
summary(improved_model)
```
```{r}
model_results <- tidy(improved_model, conf.int = TRUE) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  arrange(desc(odds_ratio))

print(model_results)
```


```{r}
probs <- predict(improved_model, newdata = data_to_model, type = "response")
preds <- as.factor(ifelse(probs > 0.5, 1, 0))

# Create the Confusion Matrix
conf_matrix <- table(Predicted = preds, Actual = data_to_model$is_cancelled)
print(conf_matrix)

# Calculate Sensitivity (Recall)
# (True Positives) / (Total Actual Positives)
recall <- conf_matrix[2,2] / sum(conf_matrix[,2])
print(paste("Recall (Cancellations caught):", round(recall, 4)))
```

```{r}
plot(improved_model, 
     which = 2, 
     main = "Q-Q Plot: Model Deviance Residuals") 
```

